Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-14 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for the choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
and both ($\beta$ and $\alpha$) show increase with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone order parameters have been observed with dehydration~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_smooth.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value at full hydration;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
  }
  	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

Lipid bilayer dehydration has been studied also with molecular dynamics simulations~\cite{mashl01,pertsin05,pertsin07,eun09,eun10,schneck12},
typically motivated by the  discussion about the origin of the ``hydration repulsion''~\cite{israelachvili,israelachvili96,sparr11}.
However, the used simulation models are typically not compared to the experimental choline and glycerol backbone
order parameters (except by Mashl et al.~\cite{mashl01}).
In Fig.~\ref{ordPhydr} the glycerol backbone and choline order parameters are shown as a function of hydration level for the CHARMM36, 
MacRog and GAFFlipid models (having the most realistic atomistic resolution structures) together with the Berger based model 
(which is the most used lipid model). 
To elucidate changes the simulation results are shifted to match better with experiments with full hydration.
The magnitude of shifting is written in the y-axis label.
The increase of the choline order parameters with dehydration is seen in all
models despite of some additional fluctuations. Thus the choline order parameter response to dehydration can be
intepreted to be in qualitative agreement with experiments.
The situation is significantly more complicated for the glycerol backbone segments 
and none of the models can be interpreted to qualitatively agree with experiments
for all the segmets.

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration (Fig.~\ref{PNangle}). However,
the amount of increase depends on the model. Especially the DLPC simulations with Berger model
predict significantly stronger P--N vector tilt compared with the other models. The Berger model
has also generally larger P--N vector angles and its choline order parameters are more off from
experiments than the other three models. Thus the relatively modest tilting with dehydration
predicted by MacRog, CHARMM36 and GAFFlipid is probably more realistic.

It should be also noted that the free energy landscape is not correct in the models which
are not able to reproduce the experimental order parameters. Thus the energetical response
to the dehydration may not be correct even though the order parameter response is qualitatively correct.
This issue may have some influence on the dehydration energetic calculations made with the Berger model~\cite{eun09,schneck12}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../Fig/dehydrationPN.pdf}

@@

The response of the glycerol backbone to dehydration seems to be more subtle than that of the choline headgroup 
and none of the models reproduced the experimental results with the accuracy allowing
the structural interpretation.

\subsubsection{Cholesterol-containing phospholipid bilayer}
Phospholipid--cholesterol interactions have been widely studied with theoretical~\cite{huang99,zhu07,rog09,alwarawrah12} and
experimental methods~\cite{brown78,marsh10,ferreira13,marsh13}, since cholesterol is abundant in biological membranes.
It has been suggested to be an important player, for example, in domain formation~\cite{simons04,somerharju09}.
It is widely agreed that cholesterol orders lipid acyl tails thus decreasing the area per molecule (condensing effect),
however, the influence of cholesterol on the lipid headgroup and glycerol backbone are sill debated~\cite{huang99,simons04,somerharju09}.
For example, it has been suggested that the surrounding lipids shield cholesterol from interactions with water by 
reorienting their headgroups (``umbrella model'')~\cite{huang99} or that cholesterol acts as a spacer for the headgroups thus increasing 
their entropy and dynamics (``superlattice model'')~\cite{somerharju09}. 
%\todo{It has been suggested that we should use direct quotations from original papers here for clarity.}
Both of these suggestions have been supported
by molecular dynamics simulations~\cite{zhu07,alwarawrah12}, and other simulations suggest specific
interactions between the glycerol backbone and cholesterol~\cite{rog09}. However,
in the above mentioned studies, the glycerol backbone and choline headgroup behaviour
as a function of cholesterol content has not been compared to experiments.

The choline headgroup and glycerol backbone order parameters for POPC measured by $^1$H-$^{13}$C NMR~\cite{ferreira13} and DPPC choline order parameters 
measured by $^{2}$H NMR~\cite{brown78} are shown in Fig.~\ref{ordPchol} as a function of cholesterol content.
The agreement between different experimental results is again very good, showing only very modest changes in 
the choline order parameters as a function of cholesterol content. It should be noted, however, that very small
changes are measurable with high resolution $^{2}$H NMR experiments
and cholesterol causes a measurable increase in the $\beta$ order parameter and a forking in the $\alpha$ order
parameter~\cite{brown78}. These effects are, however, so small that they are barely visible in the scale used in Fig.~\ref{ordPchol}.
Further, the effects of cholesterol on the glycerol backbone order parameters for POPC from $^1$H-$^{13}$C NMR experiment~\cite{ferreira13} 
are in good agreement with the results for the phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
These results further support the idea that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup stucture is independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.
    Note that to elucidate the relative change as a function of cholesterol content,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value without cholesterol;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
}
	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

In addition to the experimental data, the previously published simulation results from the Berger-POPC-07/H\"oltje-CHOL-13 model~\cite{ferreira13},
and our results from CHARMM36 and MacRog force fields are shown in Fig.~\ref{ordPchol}. 
To elucidate changes the simulation results are shifted to match better with experiments without cholesterol.
The magnitude of shifting is written in the y-axis label.
As already pointed out previosly, the Berger-POPC-07/H\"oltje-CHOL-13 model
seriously overestimates the effect of cholesterol on the phospholipid glycerol backbone and choline segments~\cite{ferreira13}.
In contrast, the responses of both CHARMM36 and MacRog are in better agreement with experiments. 
%However, CHARMM36 seems to better reproduce the experimentally observed modest changes in the glycerol backbone segments 
%g$_2$ and g$_3$ with high concentrations of cholesterol. 
Thus we have calculated the glycerol backbone dihedral angle distributions
as a function of cholesterol in CHARMM36 (shown in Supplementary material) to resolve the cholesterol induced structural changes. The only detectable change due to the
addition of cholesterol is the small decrease of gauche- and increase of gaughe+ probability of g3-g2-g1-O(\textit{sn}-1) dihedral.

>>>>

\subsubsection{Phospholipid bilayer with low hydration level}
Fig.~\ref{ordPhydr} shows the published~\cite{dvinskikh05b,ulrich94,bechinger91} experimental order parameters
for the glycerol backbone and choline as a function of hydration level.
Despite slight differences in temperature and acyl chain composition,
the three independently reported data sets for the choline ($\beta$ and $\alpha$) segments agree well with each other:
Both order parameters increase with decreasing hydration level.
The glycerol backbone order parameters (g$_{3}$, g$_{2}$, g$_{1}$), in contrast,
have been observed~\cite{dvinskikh05b} to slightly decrease with dehydration.
Note that the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91} measured only 
absolute values, but here
we included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the negative $\beta$ order parameter actually increases with dehydration 
as its absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_smooth.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value at full hydration;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
  }
  	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

Lipid bilayer dehydration has been studied also with molecular dynamics simulations~\cite{mashl01,pertsin05,pertsin07,eun09,eun10,schneck12},
typically motivated by the  discussion concerning the origin of the ``hydration repulsion''~\cite{israelachvili,israelachvili96,sparr11}.
Only one~\cite{mashl01} of these studies, however, compared their simulation model to the experimental choline and glycerol backbone
order parameters.
Fig.~\ref{ordPhydr} shows these order parameters as a function of hydration level for the CHARMM36, 
MacRog and GAFFlipid models (having the most realistic atomistic resolution structures) and a Berger-based model 
(which is the most used lipid model);
note that the simulation results have been vertically shifted to ease the comparison with experimental response to dehydration.
Despite of some fluctuations, the increase of the choline ($\beta$ and $\alpha$) order parameters is seen in all four
models. The response of the choline order parameters to dehydration can, therefore, be
interpreted to qualitatively agree with experiments.
The situation is significantly more complicated for the glycerol backbone: 
None of the four models produced the experimentally seen trends
in all the (g$_{3}$, g$_{2}$, g$_{1}$) segments.

The qualitative agreement of the $\alpha$ and $\beta$ order parameters with experiments in all four simulation models (Fig.~\ref{ordPhydr}) indicates that,
despite the unrealistic structures at full hydration (Figs.~\ref{HGorderparameters} and~\ref{FullHydrationComparisonTable}),
the structural response of the choline headgroup to dehydration is somewhat realistic.
A likely explanation is that as the interlamellar space shrinks with dehydration, the whole choline group
orients more parallel to the membrane.
Indeed, upon dehydration the angle between P--N (phoshate phosphorus to choline nitrogen) vector and membrane normal increases for
all the four models (Fig.~\ref{PNangle}). However,
the amount of increase depends on the model. Especially the DLPC simulations with Berger model
predict significantly stronger P--N vector tilt than the other models. The Berger model
has also generally larger P--N vector angles and its choline order parameters are more off from
experiments than the other three models (Fig.~\ref{HGorderparameters2}). Thus the relatively modest tilting with dehydration
predicted by MacRog, CHARMM36 and GAFFlipid is probably more realistic.

It must be stressed that in the models
incapable of reproducing the experimental order parameters the free energy landscape is not correct. Thus
even though the order parameter response to dehydration is qualitatively correct,
the energetic response is likely to be incorrect.
This may have some influence on dehydration energetic calculations made using the Berger model~\cite{eun09,schneck12}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../Fig/dehydrationPN.pdf}

@@

The response of the glycerol backbone seems to be more subtle than that of the choline headgroup;
none of the four models reproduced the experimental trends upon dehydration with enough accuracy to invite a
structural interpretation.

\subsubsection{Cholesterol-containing phospholipid bilayer}
Phospholipid--cholesterol interactions have been widely studied with theoretical~\cite{huang99,zhu07,rog09,alwarawrah12} and
experimental methods~\cite{brown78,marsh10,ferreira13,marsh13}, since cholesterol is abundant in biological membranes.
It has been suggested to be an important player, for example, in domain formation~\cite{simons04,somerharju09}.
It is widely agreed that cholesterol orders lipid acyl tails thus decreasing the area per molecule (condensing effect),
however, the influence of cholesterol on the lipid headgroup and glycerol backbone are still debated~\cite{huang99,simons04,somerharju09}.
For example, it has been suggested that the surrounding lipids shield cholesterol from interactions with water by 
reorienting their headgroups (``umbrella model'')~\cite{huang99} or that cholesterol acts as a spacer for the headgroups thus increasing 
their entropy and dynamics (``superlattice model'')~\cite{somerharju09}. 
%\todo{It has been suggested that we should use direct quotations from original papers here for clarity.}
Both of these suggestions have been supported
by molecular dynamics simulations~\cite{zhu07,alwarawrah12}, and other simulations suggest specific
interactions between the glycerol backbone and cholesterol~\cite{rog09}. However,
in the above mentioned studies, the glycerol backbone and choline headgroup behaviour
as a function of cholesterol content has not been compared to experiments.

The choline headgroup and glycerol backbone order parameters for POPC measured by $^1$H-$^{13}$C NMR~\cite{ferreira13} and DPPC choline order parameters 
measured by $^{2}$H NMR~\cite{brown78} are shown in Fig.~\ref{ordPchol} as a function of cholesterol content.
The agreement between different experimental results is again very good, showing only very modest changes in 
the choline order parameters as a function of cholesterol content. It should be noted, however, that very small
changes are measurable with high resolution $^{2}$H NMR experiments
and cholesterol causes a measurable increase in the $\beta$ order parameter and a forking in the $\alpha$ order
parameter~\cite{brown78}. These effects are, however, so small that they are barely visible on the scale used in Fig.~\ref{ordPchol}.
Further, the effects of cholesterol on the glycerol backbone order parameters for POPC from $^1$H-$^{13}$C NMR experiment~\cite{ferreira13} 
are in good agreement with the results for the phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
These results further support the idea that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup structure is largely independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.
    In order to elucidate the relative change as a function of cholesterol content,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value without cholesterol;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
}
	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

In addition to the experimental data, the previously published simulation results from the Berger-POPC-07/H\"oltje-CHOL-13 model~\cite{ferreira13},
and our results from CHARMM36 and MacRog force fields are shown in Fig.~\ref{ordPchol}. 
To elucidate changes, the simulation results are shifted to match better with experiments without cholesterol.
The magnitude of shifting is written in the y-axis label.
As already pointed out previously, the Berger-POPC-07/H\"oltje-CHOL-13 model
seriously exaggerates the effect of cholesterol on the phospholipid glycerol backbone and choline segments~\cite{ferreira13}.
In comprasion, the choline responses of both CHARMM36 and MacRog are in better qualitative agreement with experiments,
and also the glycerol backbone responses appear less drastic.
%However, CHARMM36 seems to better reproduce the experimentally observed modest changes in the glycerol backbone segments 
%g$_2$ and g$_3$ with high concentrations of cholesterol. 
Thus we have calculated the glycerol backbone dihedral angle distributions
as a function of cholesterol in CHARMM36 (shown in Supplementary material) to resolve the cholesterol-induced structural changes. The only detectable change due to the
addition of cholesterol is the small decrease of gauche(-) and increase of gaughe(+) probability of the g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral.
In conclusion, the results suggest that the significant effect of cholesterol on the lipid structures observed in simulations~\cite{zhu07,rog09,alwarawrah12} are overestimated by the model used in these studies, and the cholesterol only induces very small structural changes in the glycerol backbone.
